---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

---
title: "ase analyses of SC males"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

### Data cleaning

#### SC male DNA and RNA

aligned to hap 2 reference, reads counted using GATK ASEreadcounter

#### read in text files
```{r include=FALSE}
# required packages
library(vroom)
library(countdata)
library(tidyverse) 
library(BSDA)
library(cowplot)
library(vcfR)
library(vroom)
library(RColorBrewer)
### create repo ###
#usethis::use_git()
#usethis::use_github() 
### raw data ###
dnacountsraw <- vroom("SCF2m12dnacounts30.table", delim = "\t") # dna counts
rnacountsraw <-vroom("SCF2m12rnacounts_qual30.table", delim = "\t") #rna counts 
```

#### mutate (make variant ID), filter, combine

For (possibly predictable) reasons, the ASEReadcounter output for DNA
misses some low read-depth but confidently called heterozygous snps. This
results in some 0's in the refCount, altCount, and total count
categories in the DNA counts file - which is not expected. option -
sanity checking with the snps I have, getting depths from the vcf
directly, etc.etc

```{r combine read count files}
dnacountsraw <- mutate(dnacountsraw, variantID = str_c(contig,'_',position)) #
rnacountsraw <- mutate(rnacountsraw, variantID = str_c(contig,'_',position)) #
# cutoff of at least 16 reads - Castel et al. 2015, for RNA only
# select important columns and create ref ratio column
rnacounts <- rnacountsraw %>% filter(totalCount>=16) %>% 
  select(contig:totalCount) %>%
  mutate(rnarefratio = refCount/totalCount)
dnacounts <- dnacountsraw %>% select(contig:totalCount) %>%  filter(totalCount>=16) %>% filter(totalCount<60) %>%
  mutate(dnarefratio = refCount/totalCount) %>% filter(.,dnarefratio!=1) %>% 
  filter(.,dnarefratio!=0) 

#combocounts <- vroom("rnadnacountscombined.txt")
combocounts <- inner_join(dnacounts,rnacounts, by = c("contig","position","variantID"),
                          suffix = c("_dna","_rna")) %>%
  relocate(c(dnarefratio,rnarefratio), .after = variantID) %>% 
  mutate(filt = if_else((dnarefratio > rnarefratio & dnarefratio > 0.5 & rnarefratio > 0.5)| (dnarefratio < rnarefratio & dnarefratio < 0.5 & rnarefratio < 0.5), "toss","keep" )) %>%
  mutate(sitetype = case_when(rnarefratio > 0.5 & rnarefratio != 1  ~ "refoverexp",
                              rnarefratio < 0.5 & rnarefratio != 0 ~ "altoverexp",
                              rnarefratio == 0 ~ "refsilence", # 100% alt
                              rnarefratio == 1 ~ "altsilence", # 100% ref
                              rnarefratio == 0.5 ~ "noase")) %>% 
  filter(.,filt == "keep")

group_by(combocounts,sitetype) %>% count() # there's been no sig testing here
#vroom_write(combocounts, "rnadnacountscombined.txt",col_names=TRUE)

mean(combocounts$rnarefratio)
mean(combocounts$dnarefratio)

```
### Reference bias
mean rnarefratio is 0.570816, mean dnarefratio is 0.5432586
slightly higher than with previous genome assembly version
### site-wise statistical tests for ASE 
fisher test to see if DNA refratio is sig diff from that of RNA
binomial test for ase
```{r}
# called counts
# and then turn rna,dna, alt, ref into separate categories?
#combocounts %>% select(refCount_dna,altCount_dna,refCount_rna,altCount_rna) 
fisher <- function(a,b,c,d){
  data <- matrix(c(a,b,c,d),ncol=2)
  c(p = fisher.test(data)$p.value) #odds ratio estimate
}
bt <- function(a, b, p = 0.5) {binom.test(a, b, 0.5, alternative=
                                            c("two.sided"), conf.level = 0.95)$p.value}

# positions of sex linked regions
combocounts_sl<-combocounts %>% mutate(contig = factor(contig, levels = c("A1","A2","A4","Y1","Y2"))) %>%
  mutate(sexlink = case_when(
    (contig == 'Y1' & position > 64000000 & position < 272000000) ~ 'OldSLR',
    (contig == 'Y1' & position > 272000000) ~ 'PAR1',
    (contig == 'Y2' & position > 159000000) ~ 'OldSLR',
    (contig == 'Y1' & position < 64000000) ~ 'NewSLR',
    (contig == 'Y2' & position >121000000 & position <159000000 )  ~ 'NewSLR',
    (contig == 'Y2' & position < 121000000) ~ 'PAR2',
    contig == 'A1' ~ 'A1',
    contig == 'A2' ~ 'A2',
    contig == 'A4' ~ 'A4',
    ))

#View(select(pvals, refCount_rna,altCount_rna, refCount_dna,altCount_dna,p))
gc()

pvals <- combocounts_sl %>%
  rowwise() %>%
  mutate(pfisher=fisher(refCount_dna,altCount_dna,refCount_rna,altCount_rna)[[1]]) %>%
  mutate(pbinom=bt(refCount_rna,totalCount_rna)) #%>% ## FDR?
pvals<- pvals %>% mutate(.,pfisherBH = (p.adjust(pfisher,"BH",n=length(pvals$pfisher)))) %>%
  mutate(.,pbinomBH = (p.adjust(pbinom,"BH",n=length(pvals$pbinom)))) 
#apply benjaminin hochberg correction R
nrow(filter(pvals, pfisherBH<0.05,pbinomBH<0.05)) # 6986
nrow(filter(pvals, pbinomBH<0.05)) #22878 

sigtest<- pvals %>% filter(pfisherBH<0.05) %>% filter(pbinomBH<0.05) %>% 
  group_by(contig,sitetype) %>% summarise(n_sig = n())
#just_binom <- pvals %>% filter(pbinomBH<0.05) %>% 
#  group_by(contig,sitetype) %>% summarise(n_sig = n())
totalnum <- pvals %>% group_by(contig) %>% summarise(n_all= n())
#mutate(propn, prop = n_sig/n_all) 
#totalnum <- group_by(snplist, CHROM) %>% summarise(n=n()) %>% arrange(.,desc(n))
propn<- left_join(sigtest, totalnum, by = c("contig")) %>% mutate(., prop = n_sig/n_all)
#vroom_write(pvals, "sc_male_ase_sigtests.csv",",")
# fisher and binom test
sigtest_sl<- pvals %>% filter(pfisherBH<0.05) %>% filter(pbinomBH<0.05) %>% 
  group_by(sexlink,sitetype) %>% summarise(n_sig = n())
totalnum_sl <- pvals %>% group_by(sexlink) %>% summarise(n_all= n())
propn_sl <- left_join(sigtest_sl,totalnum_sl, by = c("sexlink")) %>% 
  mutate(., prop = n_sig/n_all)
print(propn_sl)
## just the binom test
sigtest_binom<- pvals %>% filter(pbinomBH<0.05) %>% 
  group_by(sexlink,sitetype) %>% summarise(n_sig = n())
totalnum_sl <- pvals %>% group_by(sexlink) %>% summarise(n_all= n())
propn_binom <- left_join(sigtest_binom,totalnum_sl, by = c("sexlink")) %>% 
  mutate(., prop = n_sig/n_all)
#vroom_write(pvals, "sigtable.txt")
```
### beta-binomial test?
```{r}
#countdata::bb.test()

#bb <- function(a, b, p = 0.5) {bb.test(a, b, 0.5, alternative=
#                                            c("two.sided"))$p.value}
#pvals_beta <- combocounts_sl %>%
#  rowwise() %>%
  #mutate(pfisher=fisher(refCount_dna,altCount_dna,refCount_rna,altCount_rna)[[1]]) %>%
#  mutate(pbinom=bb(refCount_rna,totalCount_rna)) #%>% ## FDR?
```

### Analyses
#### to do
-   sig snp in gene = gene has sig ASE - count only once
-   repeat w hap1, use liftover/ortho list
-   later - snps, anno TEs, connect to genes, etc. -e..g median/mean
    across genes or something This table is just the number of ASE sites
    where there is significant departure b/w the dna and rna w respect
    to alt and ref read counts. Need to filter a bit more probably.
    Plotting ASE vs contig to get a better idea of whats going on here.
-   denom for propn should be snps after filtering but before sign \####
    getting total counts for proportions of sig sites

```{r}
#myvcf<-read.vcfR("scf2m12_dna_noindelsqual60_hap2.recode.vcf.gz")
#snpnum<-myvcf@fix %>% group_by(CHROM) %>% summarise()
#myvcfdf <- as.data.frame(myvcf@fix)
#rm(myvcf)
#head(myvcfdf)
#snplist<-select(myvcfdf, CHROM,POS)
#vroom_write(snplist, "snplist.txt", delim = "\t", col_names = TRUE)
#snplist<-vroom("snplist.txt")
#rm(snplisttest)
#group_by(snplist, CHROM) %>% summarise(n=n()) %>% arrange(.,desc(n))
```

#### exploratory plots

DNA read count distributions

* Settled on 60 as a max read count.

```{r}
## plot total DNA counts to see if a "max" read count is useful at all
# ggplot(pvals, aes(totalCount_dna)) + geom_histogram(binwidth=5) + xlim(0,60) # removes 867 rows
#ggplot(pvals, aes(totalCount_dna)) + geom_histogram(binwidth=5)
# md <-mean(pvals$totalCount_dna)
# mean(pvals$totalCount_dna)
# median(pvals$totalCount_dna)
# max(pvals$totalCount_dna)
# min(pvals$totalCount_dna)
# 
# # https://academic.oup.com/bioinformatics/article/30/20/2843/2422145
# # Each chosen threshold yields a point in the plot. An arrow points to a point on the MD curve when the corresponding read depth is right above d+4d‾‾√ , where d is the mean read depth across called variants
# #maxdepth <- md + (4*sqrt(md))
# #maxdepth # 51
# # not sure i like this cutoff... much to think about
# ## orr... I should go to the vcf maybe
# #"Usually a good rule of thumb is something the mean depth x 2 - so in this case we could set our maximum depth at 40x." - https://speciationgenomics.github.io/filtering_vcfs/
# maxdepth <- md*2
# maxdepth # 59? - maybe try from the vcf... 
# #md 60
# 
# ggplot(pvals_md, aes(totalCount_dna)) + geom_histogram(binwidth=5) + xlim(0,60) # removes 867 rows
# ggplot(pvals_md, aes(totalCount_dna)) + geom_histogram(binwidth=5)
# #md <-mean(pvals$totalCount_dna)
# mean(pvals_md$totalCount_dna)
# median(pvals_md$totalCount_dna)
# max(pvals_md$totalCount_dna)
# min(pvals_md$totalCount_dna)
# 
# propn<- left_join(sigtest, totalnum, by = c("contig"="contig")) %>% mutate(., prop = n_sig/n_all)
```

RNA ref ratio

* not filtered by significance
* lower ratios in Y1, reference bias everywhere else, no significant diff b/w groups

```{r}
#ggplot(combocounts, aes(contig, rnarefratio)) + geom_col() +theme_bw() + 
#  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) # this plot?? $ plot the mean!!!! or median!

#ggplot(combocounts, aes(contig, dnarefratio)) + geom_col() +theme_bw() +
#  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ase_df<- pvals %>% #filter(pfisherBH<0.05 & pbinomBH<0.05)
   group_by(contig) %>% 
    summarise(mean_ase = mean(rnarefratio),
  sd_ase=sd(rnarefratio))
ggplot(ase_df)

 ase_rna<-ggplot(ase_df,aes(contig, mean_ase)) +
    geom_col(width = 0.85) +
    geom_errorbar(aes(ymin = mean_ase - sd_ase,
                      ymax = mean_ase + sd_ase),
                  color = "#22292F",
                  width = .1) +
   theme_classic() +  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
   ylab("mean reference:total read count ratio (RNA)") + geom_hline(yintercept=0.5, linetype="dashed") + ylim(0,1) + ggtitle("RNA ref:total count")
#ggsave("asemean_all_rna.png", height=4,width=7)

 ase_df_dna<- pvals %>% #filter(pfisherBH<0.05 & pbinomBH<0.05)
   group_by(contig) %>% 
    summarise(mean_ase = mean(dnarefratio),
  sd_ase=sd(dnarefratio))
ase_dna<- ggplot(ase_df_dna,aes(contig, mean_ase)) +
    geom_col(width = 0.85) +
    geom_errorbar(aes(ymin = mean_ase - sd_ase,
                      ymax = mean_ase + sd_ase),
                  color = "#22292F",
                  width = .1) +
   theme_classic() +  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
   ylab("mean reference:total read count ratio (DNA)") +geom_hline(yintercept=0.5, linetype="dashed") +
  ggtitle("DNA ref:total count") +ylim(0,1)
#ggsave("asemean_all_dna.png", height=5,width=7)
cowplot::plot_grid(ase_rna,ase_dna)
```
RNA vs DNA refratio - by sex linked region

* Old Sex linked region has most obvious dip in mean ase 
```{r}
ase_df_sl<- pvals %>% #filter(pfisherBH<0.05 & pbinomBH<0.05)
   group_by(sexlink) %>% 
    summarise(mean_ase = mean(rnarefratio),
  sd_ase=sd(rnarefratio))

rna_ase_sl<-ggplot(ase_df_sl,aes(sexlink, mean_ase)) +
    geom_col(width = 0.85) +
    geom_errorbar(aes(ymin = mean_ase - sd_ase,
                      ymax = mean_ase + sd_ase),
                  color = "#22292F",
                  width = .1) +
   theme_classic() +  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
   ylab("mean reference:total read count ratio (RNA)") + geom_hline(yintercept=0.5, linetype="dashed") + ylim(0,1)
#ggsave("asemean_all_rna_sex.png", height=5,width=7)

 ase_df_dna_sl<- pvals %>% #filter(pfisherBH<0.05 & pbinomBH<0.05)
   group_by(sexlink) %>% 
    summarise(mean_ase = mean(dnarefratio),
  sd_ase=sd(dnarefratio))
dna_ase_sl<-ggplot(ase_df_dna_sl,aes(sexlink, mean_ase)) +
    geom_col(width = 0.85) +
    geom_errorbar(aes(ymin = mean_ase - sd_ase,
                      ymax = mean_ase + sd_ase),
                  color = "#22292F",
                  width = .1) +
   theme_classic() +  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
   ylab("mean reference:total read count ratio (DNA)") + geom_hline(yintercept=0.5, linetype="dashed") + ylim(0,1)
#ggsave("asemean_all_dna_sex.png", height=5,width=7)
cowplot::plot_grid(rna_ase_sl,dna_ase_sl)

```

#### plot signifiance ase proportions per chromosome/region

By chromosome

* Y1 and Y2 have elevated ref (Y) silencing, alt (X) overexpression
```{r}
ggplot(propn, aes(contig,prop,fill=sitetype)) + 
  geom_col() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  ylab("proportion of sites with significant ASE") #+
#ggsave("propnsigASE_xyymale_fixed.png", width = 7, height = 5)

```
By sex-linked region (as defined in genome paper draft)

* Old sex-linked region has an extreme amount of Y silencing
* changing p value cut off to 0.1 doesn't change the pattern much
```{r}
#ggplot(filter(propn_sl,contig != 'Scaffold_71'), aes(contig,prop,fill=sitetype, group=sexlink)) + #geom_col(position="dodge",aes(group=contig)) + 
 #  geom_bar(stat = prop,
   #        position = "dodge") +
  #theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + theme_classic() 
ggplot(filter(propn_sl,!is.na(sexlink)), aes(sexlink,prop,fill=sitetype)) + 
  geom_col() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  ylab("proportion of sites with significant ASE") #+ scale_fill_discrete(name = "Site type", labels = c("X overexpressed","X silenced","Y overexpressed","Y silenced")) 
  
#ggsave("propnsigASE_sexlinkage_xyymale_fixed.png", width = 7, height = 5)
#ggsave("propnsigASE_xyymale_2022_11_22.png", width = 7, height = 5)

## stick into google doc
## genes - go analysis
# how it corresponds to fraction of genes
# patterns across genome - density of Y-silenced across etc.

sigtest_sl_01<- pvals %>% filter(pfisherBH<0.1) %>% filter(pbinomBH<0.1) %>% 
  group_by(sexlink,sitetype) %>% summarise(n_sig = n())
totalnum_sl <- pvals %>% group_by(sexlink) %>% summarise(n_all= n())
propn_sl_01 <- left_join(sigtest_sl_01,totalnum_sl, by = c("sexlink")) %>% 
  mutate(., prop = n_sig/n_all)

ggplot(filter(propn_sl_01,!is.na(sexlink)), aes(sexlink,prop,fill=sitetype)) + 
  geom_col() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  ylab("proportion of sites with significant ASE")
ggsave("propnsigASE_xyymale_2023_05_22.png", width = 7, height = 5)

```

Plot position by ASE
```{r}
y1plot<-ggplot(filter(pvals,contig=="Y1"), aes(position,rnarefratio)) +geom_point(aes(color=sexlink)) +geom_smooth() +theme_classic()

y2plot<-ggplot(filter(pvals,contig=="Y2"), aes(position,rnarefratio)) +geom_point(aes(color=sexlink)) +geom_smooth() +theme_classic() 
cowplot::plot_grid(y1plot,y2plot)
#legend1 <- get_legend(
  # create some space to the left of the legend
#  y1plot + theme(legend.box.margin = margin(0, 0, 0, 12))
#)
##
y1plotsig<-ggplot(filter(pvals,contig=="Y1",pbinomBH<0.1,pfisherBH<0.1), aes(position,rnarefratio)) +geom_point(aes(color=sexlink)) +geom_smooth() +theme_classic()

y2plotsig<-ggplot(filter(pvals,contig=="Y2",pbinomBH<0.1,pfisherBH<0.1), aes(position,rnarefratio)) +geom_point(aes(color=sexlink)) +geom_smooth() +theme_classic() 
ggsave("y1sigplot.png",plot=y1plotsig, width=7, height=5)
ggsave("y2sigplot.png",plot=y2plotsig, width=7, height=5)
cowplot::plot_grid(y1plotsig,y2plotsig)
##
ggplot(filter(pvals,pbinomBH<0.1,pfisherBH<0.1,contig=="Y1"), aes(x=position)) +geom_histogram(bins=50,aes(fill=sexlink))

ggplot(filter(pvals,pbinomBH<0.1,pfisherBH<0.1,contig=="Y2"), aes(x=position)) +geom_histogram(bins=50,aes(fill=sexlink)) 

ggplot(filter(pvals,contig=="Y1"), aes(x=position)) +geom_histogram(bins=50,aes(fill=sexlink)) 

ggplot(filter(pvals,contig=="Y2"), aes(x=position)) +geom_histogram(bins=50,aes(fill=sexlink)) 
```

```{r eval=FALSE, include=FALSE}
# ggplot(filter(propn_binom,!is.na(sexlink)), aes(sexlink,prop,fill=sitetype)) + 
#   geom_col() +
#   theme_classic() +
#   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
#   ylab("proportion of sites with significant ASE\n (binomial test only)") +
#   scale_fill_discrete(name = "Site type", labels = c("X overexpressed","X silenced","Y overexpressed","Y silenced"))
# #ggsave("propnsigASE_binomonly_xyymale_2022_11_22.png", width = 7, height = 5)
```

```{r eval=FALSE, include=FALSE}
#### What genes are these snps in anyway??
# did some bedtools/awk stuff to get everything in the right shape first
# aight time to merge ASE results to my quasi-bed file. scaf_pos for this new file created by concatenating the scaffold and stop positions (since that's equivalent to snp pos in a vcf. hope that works?)
# anno<-vroom("scmale_snpgeneintersect_scafpos.txt", col_names=c("contig","start","stop","variantID","contig_gene","start_gene","stop_gene","geneID","annotation","feature"))
# #looks as it should. variantID matches exist!
# #genes<- filter(anno, feature == "gene")
# unique genes? 
#n_distinct(genes$variantID) 
# ase_anno<-inner_join(pvals,anno, by = c("variantID"))
# genes_per_site<- ase_anno %>% filter(feature == "gene") %>% 
#   group_by(variantID) %>% summarise(mean(rnarefratio), n()) # did # overlapping genes are just on diff strands its all gucci
# # snps to be handled the same way. i think
# # lol
# site_per_gene<- ase_anno %>% filter(feature == "gene") %>% 
#   group_by(geneID,sexlink) %>% summarise(dnarefratio=mean(dnarefratio),rnarefratio=mean(rnarefratio), number=n())  # there's a gene w 100's of snps in it lmao
# ## sig
# site_per_gene_sig<- ase_anno %>% filter(feature == "gene", pbinomBH <0.05, pfisherBH<0.05) %>% 
#   group_by(geneID) %>% summarise(meanrefratio=mean(rnarefratio), meandnaratio=mean(dnarefratio), number=n())
# write_csv(ase_anno,"ase_snp_annotation_combined.csv",col_names = T)  
```

```{r eval=FALSE, include=FALSE}
# ggplot(site_per_gene,aes(rnarefratio)) + geom_histogram() + facet_wrap(~sexlink)
# ggplot(site_per_gene,aes(dnarefratio)) + geom_histogram() + facet_wrap(~sexlink)

```
### to do
* try plotting density of y-silenced snps
  * clustering?
* fraction of silenced (sig underexp) snps differ b/w sal origins??
* epigenetic silencing??? e.g. TEs - strata or TE lol
* 
 
